---
title: "Simple Regression"
subtitle: "Elections and the Economy"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NULL)
```

In this tutorial I will use the following R packages:

```{r, warning=FALSE, message=FALSE}
library(car)
library(arm)
library(psych)
```

This example is taken from Gelman, Hill, and Vehtari (2021), which is taken from the bread and peace model by Douglas Hibbs.
This model is based on two factors that dominate the election of U.S. presidents post World War II.
These two factors are weighted-average growth per capita disposable income over the prior term, and the number of U.S. fatalities resulting from unprovoked deployment.
This example will focus on the first factor.

```{r}
source("code/importCleanHibbs.R")
```


```{r}
scatterHist(hibbs[2:3], xlim = c(-5, 8), ylim = c(35, 70), smooth = FALSE)
```


```{r}
describe(hibbs[2:3], fast = TRUE)
```


```{r}
source("code/breadPeaceModelPlot.R")
```


```{r}
mod <- lm(vote ~ growth, data = hibbs)
display(mod)
```


```{r}
# simulate model ----------------------------------------------------------
simHibbs <- function(model) {
  a <- coef(model)[1]
  b <- coef(model)[2]
  x <- model$model[[2]]
  n <- length(x)
  sigma <- summary(mod)$sigma
  simvote <- a + b*x + rnorm(n, 0, sigma)
  simmod <- lm(simvote ~ x)
  return(c(a = coef(simmod)[1], b = coef(simmod)[2], sigma = summary(simmod)$sigma))
  }

sims <- replicate(10000, simHibbs(mod))
sims <- t(sims)
head(sim)
```


```{r}
plot(vote ~ growth, data = hibbs, xlim = c(-2, 6), ylim = c(40, 70), type = "n")
#idx <- sample(1:nrow(sims), 100)
for(i in 1:nrow(sims)) {
  abline(a = sims[i, 1], b = sims[i, 2], col = "grey")
}
points(x = hibbs$growth, y = hibbs$vote)
abline(coef(mod), col = "black")
abline(h = 50, lty = 2, col = "darkblue")
```


